---
title: "Assignment 2: Coding Basics"
author: "Ardath Dixon"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. 
ListByFourTo100 <- seq(1,100,4) #creates sequence and assign name
ListByFourTo100 #shows values in console

#2. 
ListByFour_mean <- mean(ListByFourTo100) #assigns name to mean
ListByFour_median <- median(ListByFourTo100) #assigns name to median

ListByFour_mean #shows mean value in console
ListByFour_median #shows median value in console

#3. 
ListByFour_mean > ListByFour_median #gives FALSE output because they're equal

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5.
StudentScores <- c("Abe"=85, "Bobbi"=93, "Carly"=82, "Dan"=45) #numeric vector


PassingGrade <- function(x){
  passed <- x>49
  return(list(passed))
}
PassingGrade(StudentScores)

#7.
df_StudentScores <- as.data.frame(StudentScores)

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: This data frame lists vectors of different modes, while a matrix requires all columns to have the same mode. 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: 
