---
title: "Assignment 4: Data Wrangling"
author: "Ardath Dixon"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A04_DataWrangling.Rmd") prior to submission.

The completed exercise is due on Tuesday, Feb 16 @ 11:59pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1
getwd()
setwd("/Users/ardat/OneDrive/Documents/DataAnalytics/Environmental_Data_Analytics_2021/")

#install.packages("tidyverse") ## run line if tidyverse is not yet installed on your machine
#install.packages("lubridate") ## run line if lubridate is not yet installed on your machine
library(tidyverse)
library(lubridate)

EPAair_O3_NC2018 <- read.csv('./Data/Raw/EPAair_O3_NC2018_raw.csv', stringsAsFactors = TRUE)
EPAair_O3_NC2019 <- read.csv('./Data/Raw/EPAair_O3_NC2019_raw.csv', stringsAsFactors = TRUE)
EPAair_PM25_NC2018 <- read.csv('./Data/Raw/EPAair_PM25_NC2018_raw.csv', stringsAsFactors = TRUE)
EPAair_PM25_NC2019 <- read.csv('./Data/Raw/EPAair_PM25_NC2019_raw.csv', stringsAsFactors = TRUE)

#2
dim(EPAair_O3_NC2018)
colnames(EPAair_O3_NC2018)

dim(EPAair_O3_NC2019)
colnames(EPAair_O3_NC2019)

dim(EPAair_PM25_NC2018)
colnames(EPAair_PM25_NC2018)

dim(EPAair_PM25_NC2019)
colnames(EPAair_PM25_NC2019)
```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3 Change date from character to date format

## When datasets were imported without the 'stringsAsFactors' command above, the format commands below were not necessary. Once they were imported as Factors, the code produced error messages until the format commands were included.

EPAair_O3_NC2018$Date <- format(EPAair_O3_NC2018$Date, format="%m/%d/%Y")
EPAair_O3_NC2018$Date <- as_date(EPAair_O3_NC2018$Date, format="%m/%d/%Y")

EPAair_O3_NC2019$Date <- format(EPAair_O3_NC2019$Date, format="%m/%d/%Y")
EPAair_O3_NC2019$Date <- as_date(EPAair_O3_NC2019$Date, format="%m/%d/%Y")

EPAair_PM25_NC2018$Date <- format(EPAair_PM25_NC2018$Date, format="%m/%d/%Y")
EPAair_PM25_NC2018$Date <- as_date(EPAair_PM25_NC2018$Date, format="%m/%d/%Y")

EPAair_PM25_NC2019$Date <- format(EPAair_PM25_NC2019$Date, format="%m/%d/%Y")
EPAair_PM25_NC2019$Date <- as_date(EPAair_PM25_NC2019$Date, format="%m/%d/%Y")

#4 Select certain columns

Processed_EPAO3_NC2018 <- select(EPAair_O3_NC2018, Date, DAILY_AQI_VALUE:Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
Processed_EPAO3_NC2019 <- select(EPAair_O3_NC2019, Date, DAILY_AQI_VALUE:Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
Processed_PM25_NC2018 <- select(EPAair_PM25_NC2018, Date, DAILY_AQI_VALUE:Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)
Processed_PM25_NC2019 <- select(EPAair_PM25_NC2019, Date, DAILY_AQI_VALUE:Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

#5 Fill in PM2.5 for all AQS_PARAMETER_DESC cells in the PM2.5 datasets

Processed_PM25_NC2018 <- mutate(Processed_PM25_NC2018, AQS_PARAMETER_DESC = "PM2.5")
Processed_PM25_NC2019 <- mutate(Processed_PM25_NC2019, AQS_PARAMETER_DESC = "PM2.5")

#6 Save all processed datasets as .csv files

write.csv(Processed_EPAO3_NC2018, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")
write.csv(Processed_EPAO3_NC2019, row.names = FALSE, file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(Processed_PM25_NC2018, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(Processed_PM25_NC2018, row.names = FALSE, file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1718_Processed.csv"

```{r}
#7 Combine the 4 datasets into 1

EPAair_O3_PM25_NC201819 <- rbind(Processed_EPAO3_NC2018, Processed_EPAO3_NC2019, Processed_PM25_NC2018, Processed_PM25_NC2019)

#8 Include specific sites, generate daily means, add month & year columns

EPAair_O3_PM25_NC201819_commonsites <-
  EPAair_O3_PM25_NC201819 %>%
  filter(Site.Name == "Linville Falls" | Site.Name == "Durham Armory" | Site.Name == "Leggett" | Site.Name == "Hattie Avenue" | Site.Name == "Clemmons Middle" | Site.Name == "Mendenhall School" | Site.Name == "Frying Pan Mountain" | Site.Name == "West Johnston Co." | Site.Name == "Garinger High School" | Site.Name == "Castle Hayne" | Site.Name == "Pitt Agri. Center" | Site.Name == "Bryson City" | Site.Name == "Millbrook School") %>%
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY) %>%
  summarise(meanAQI = mean(DAILY_AQI_VALUE),
            meanLat = mean(SITE_LATITUDE),
            meanLong = mean(SITE_LONGITUDE)) %>%
  mutate(Month = month(Date), Year = year(Date))
  
#9 Spread AQI values into two columns per site: ozone & PM2.5

EPAair_O3_PM25_NC201819_tidy <-
  EPAair_O3_PM25_NC201819_commonsites %>%
  pivot_wider(names_from = AQS_PARAMETER_DESC, values_from = meanAQI)

#10 See dimensions

dim(EPAair_O3_PM25_NC201819_tidy)

#11 Save as a .csv -- instructions said 1718 but this data is 1819, so I changed the name accordingly

write.csv(EPAair_O3_PM25_NC201819_tidy, row.names = FALSE, file = "./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a

EPAair_O3PM25_1819_Summary <-
  EPAair_O3_PM25_NC201819_tidy %>%
  group_by(Site.Name, Month, Year) %>%
  summarise(meanAQI_ozone = mean(Ozone),
            meanAQI_PM2.5 = mean(PM2.5))

#12b

EPAair_O3PM25_1819_dropna <- 
  pivot_longer(EPAair_O3PM25_1819_Summary, meanAQI_ozone:meanAQI_PM2.5, 
               names_to = "AQS_PARAMETER_DESC", values_to = "meanAQI")  %>%
  group_by(Site.Name, Month, Year) %>%
  drop_na(meanAQI)

#13

dim(EPAair_O3PM25_1819_dropna)
```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: We used drop_na rather than na.omit because drop_na actually changes the dataframe. The other function (na.omit) only takes away the na values for that line of processing without making broader changes. 
