---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Ardath Dixon"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd() ## setwd() if need to change

library(tidyverse)
library(agricolae)
library(lubridate)

NTL_chemphys <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
NTL_chemphys$sampledate <- as.Date(NTL_chemphys$sampledate, format = "%m/%d/%y")

#2
mytheme <- theme_bw(base_size = 10)+
  theme(axis.text = element_text(color="black"))

theme_set(mytheme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Lake temperatures recorded in July have no variation among different depths.
Ha: Lake temperatures recorded in July vary among different depths.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

NTLchemphys_month <- mutate(NTL_chemphys, month = month(sampledate))

Julydepthtemp <- NTLchemphys_month %>%
  filter(month == 7) %>%  
  select(lakename:daynum, depth:temperature_C) %>%
  drop_na()
  
#5
Julytempdepth_scatter <- ggplot(Julydepthtemp, aes(x=depth, y=temperature_C))+
  geom_point(color = "cadetblue2")+
  geom_smooth(method=lm, color = "black")+
  ylim(0,35)+
  theme_dark()+
  labs(x = "Depth below Water Surface (m)", y = "Temperature (C)")
print(Julytempdepth_scatter)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: This figure suggests that the deeper underwater a measurement was taken, the colder the recorded temperature. The points have more range at shallower depth and less of a range at deeper depths. This suggests that this trend could be more of a curve than a linear relationship. This could be due to shady or sunny patches at the surface, where surface temperature is more vulnerable to its surroundings than deeper measurements. Meanwhile, the temperature has clear limits and can not increase or decrease exponentially to infinity, so a cubic root function would be more likely. 



7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
Julydepthtemp.regression <- lm(data = Julydepthtemp, temperature_C~depth)
summary(Julydepthtemp.regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: This simple linear regression suggests a significant difference between temperatures recorded at different depths, since the p-value < 2.2e-16. It suggests that depth changes account for 73.87% of the variability in temperature. This is based on 9726 degrees of freedom (9728 observations - 2 variables compared). This predicts that every 1m change in depth will result in 1.946 degrees lowered temperature.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9

Julytempdepth_AIC <- lm(data = Julydepthtemp, temperature_C ~ depth + year4 + daynum)
step(Julytempdepth_AIC)

## All three tested variables (depth, year, daynum) show higher AIC values than the <none> 
## control, and therefore should be included for multiple regression analysis.

#10

Julytempdepth_multreg <- lm(data = Julydepthtemp, temperature_C ~ depth + year4 + daynum)
summary(Julytempdepth_multreg)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests including all variables tested as contributing factors for temperatures -- depth, year of recording, and that season's day of recording. Analyses without these variables each result in higher AIC values than the <none> control, and since lower AIC is preferable, including them can provide an improved analysis. These factors explain 0.741 of the variance (74.1%), which is an improvement from 0.739 when depth was the only explanatory variable.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
Julydepthtemp_ANOVA <- aov(data=Julydepthtemp, temperature_C ~ lakename)
summary(Julydepthtemp_ANOVA)

Julydepthtemp_ANOVA2 <- lm(data=Julydepthtemp, temperature_C ~ lakename)
summary(Julydepthtemp_ANOVA2)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, the ANOVA tests shows a significant difference in mean temperature among lakes, since the ANOVA test gives p<2e-16. Certain lakes show more extreme mean temperature differences shown in the range of p-values from the lm test, yet all are < 0.05, so therefore confirm the conclusion to recognize significant differences in mean temperature among the lakes. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
Lakes_tempdepth_scatter <- ggplot(Julydepthtemp, aes(x=depth, y=temperature_C, 
                                                     color = lakename))+
  geom_point(alpha = 0.5)+
  geom_smooth(method=lm, se = FALSE)+
  ylim(0,35)+
  labs(x = "Depth below Water Surface (m)", y = "Temperature (C)", color = "Lake Name")+
  theme_gray()
print(Lakes_tempdepth_scatter)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(Julydepthtemp_ANOVA)

Lakes_tempdepth <- HSD.test(Julydepthtemp_ANOVA, "lakename", group = TRUE)
Lakes_tempdepth

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Statistically speaking, Peter has the same mean temperature as Paul Lake (p = 0.224) and Ward Lake (p = 0.783). No lake has a statistically distinct mean temperature from all other lakes. This is shown in the groups of the HSD.test, where no letter is assigned to solely one lake (e.g. Central Long Lake and Crampton Lake both belong to group a).

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: Looking soley at Peter Lake and Paul Lake for mean temperature comparisons, the T-test could show if they have distinct mean temperatures. The T-Test can be used for a continuous response variable (temperature) from one categorical variable with two categories (Peter Lake vs. Paul Lake), so fits well in this application.


